---
title: " Raster data"
output: 
  html_document: 
    theme: yeti
---

```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE}
## This chunk automatically generates a text .R version of this script when running within knitr.  You do not need to run this...
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
knitr::purl(input,output,documentation=2,quiet=T,envir=new.env())
knitr::opts_chunk$set(cache=TRUE)
```

[<i class="fa fa-file-code-o fa-3x" aria-hidden="true"></i> The R Script associated with this page is available here](`r output`).  Download this file and open it (or copy-paste into a new script) with RStudio if you want to follow along.  

# Working with raster data

Raster data in R can be loaded in a relatively straight forward way using the **raster** package. There are other ways of loading raster data, including implementations in **sp**, where raster data can be formatted as 'SpatialPixelsDataFrame()'. 

The **raster** package is still the most widely used package though, although it is increasingly being replaced by an approach that is based on a tidy concept (separating raw data from metadata and spatial information). More on this later

First, lets load the raster package and some first data

```{r}
library(raster)

# The data to load is from Jung et al. (2020), https://www.nature.com/articles/s41597-020-00599-8
# I have prepared a subset for the Alpes to load in here. 
# These contain the fraction (multiplied with 1000) of a grid cell containing forest or arti at 1km. 

# We now load the forest layer from this object
ras <- raster('ht_004_clipped.tif',layer = 1)

class(ras)

# Plot the raster data
plot(ras, col = topo.colors(100))

```

Here we loaded in the first band of the data as Raster object, but there are others (more about this in [multidimensional](multidim.html)).

<div class="well">
Can you load in and plot the second band (artifical landscapes)?

<button data-toggle="collapse" class="btn btn-primary btn-sm round" data-target="#demo1">Show Solution</button>
<div id="demo1" class="collapse">
```{r, purl=F}
ras_crop <- raster('ht_004_clipped.tif',layer = 2)

plot(ras, col = rev(heat.colors(100)))

```

</div>
</div>

Statistics and data information



