---
title: "Extractions and overlays"
output: 
  html_document: 
    theme: yeti
---

```{r, echo=FALSE, message=FALSE, results='hide', purl=FALSE}
## This chunk automatically generates a text .R version of this script when running within knitr.  You do not need to run this...
input  = knitr::current_input()  # filename of input document
output = paste(tools::file_path_sans_ext(input), 'R', sep = '.')
knitr::purl(input,output,documentation=2,quiet=T,envir=new.env())
knitr::opts_chunk$set(cache=TRUE)
```

[<i class="fa fa-file-code-o fa-3x" aria-hidden="true"></i> The R Script associated with this page is available here](`r output`).  Download this file and open it (or copy-paste into a new script) with RStudio if you want to follow along.  

# Raster extractions

Data extractions are quite common GIS tasks both for vector and raster data. Here the raster package support a number of functions to extract values from another or different data sources (vector files).

It is possible to extract data directly from a raster file
```{r, eval =FALSE}
library(raster)

# Get the forest cover
ras <- raster('ht_004_clipped.tif', 1)
ras <- setMinMax(ras)

# The value in the upper left corner
ras[1,1]

# Take 10 point values at random
sampleRandom(ras, 10)

# Or regularly spaced
sampleRegular(ras, 10 )

# or stratefied in case you have zones 
# sampleStratified()

```

Note that the **sf** package has similar methods, see `st_sample()`

Another common application are 'zonal' statistics, so where one raster layer provides the target estimates to be extracted and another specific zones for which the values are requested.

```{r}
# Load both forest and artifical land cover data
forest <- raster('ht_004_clipped.tif',1)
artif <- raster('ht_004_clipped.tif',2)

# Create a zone based on artifical coverage
# E.g. values below 50% get 1 and above get 2
artif <- reclassify(artif, c(0,500,1,
                             500,1000,2)
                    )
# Check that both have the same extent and resolution, otherwise need to rescale
compareRaster(forest,artif)

# Now calculate for both zones the mean fraction of forest within them 
zonal(forest, artif, fun = 'mean')

```

Considerable more common however is the extraction of rasterdata within a given polygon or point

```{r}
# Artifical
ras <- raster('ht_004_clipped.tif',2)
# Load Laxenburg shape
laxpol <- st_read('Laxenburg.gpkg')

ex <- extract(ras, laxpol, fun = mean)
ex

# Another useful way for extracting values is to first clip and mask the input raster
# Often this can be considerably faster speedwise
# First crop
ras <- crop(ras, laxpol)
# Then mask
ras <- mask(ras, laxpol)
# Note the options for masking though (?mask)

# Now we simply get all values and calculate the mean directly
mean( values(ras) , na.rm = T)

```


## Even faster extractions

There are other packages such as ´velox´ or ´exactextractr´ that support even faster extraction of raster data.
Here we extract the coverage fraction of each

```{r}
library(sf)
library(raster)
library(exactextractr)
# Load Forest
ras <- raster('ht_004_clipped.tif', 1)

# Load Laxenburg polygon
laxpol <- st_read('Laxenburg.gpkg')

# The coverage fraction of each value
frac  <- exact_extract(ras, laxpol)[[1]]
frac

# Or alternatively summarize a mean
exact_extract(ras, laxpol, fun = 'mean')
```

